OPP Veil of Darkness Report

### 2018-12-13 lib/vod/scraps/vodnotebook.Rmd
cleaned up vod into lib/vod

```{r}
data <- read_rds(here::here("cache", "vod_full_data.Rds"))
```

@@

```{r}
plot_prop_minority_by_time(
    data,
    city_only = "Madison",
    title = "Madison, WI",
    min_clock_time = hms::hms(hours = 18, min = 30), 
    max_clock_time = hms::hms(hours = 19, min = 30),
    smooth_method = "lm"
  )
```

@@

@@

by_city_dst %>%  
  mutate(low = estimate - 1.96*std.error, high = estimate + 1.96*std.error) %>% 
  ggplot(aes(reorder(city, estimate), estimate)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_point() +
  geom_errorbar(aes(ymin = low, ymax = high)) +
  theme(
    axis.text.x = element_text(hjust = 1, angle = 30)
  ) +
  labs(
    x = NULL,
    y = "Coefficient on isDarkTRUE with 95% conf int",
    title = "Clock-time and Subgeography Adjusted DST Models by City"
  ) +
  facet_grid(season ~ .)
```

@@

plot_data <-
  full_data %>% 
  filter(
    days_since_dst < days(15),
    days_since_dst > days(-15),
    minute >= min_minute,
    minute <= min_minute + 20,
    city != "El Paso"
  ) %>% 
  mutate(
    is_dark = factor(is_dark, levels = c(TRUE, FALSE), labels = c("Dark", "Light")),
    before_dst = days_since_dst < 0
  ) %>% 
  group_by(days_since_dst, before_dst, season, is_dark, geo_control) %>% 
  summarise(prop_minority = sum(is_minority_demographic) / n(), n = n()) %>%
  summarise(weighted_mean_prop = weighted.mean(prop_minority, n))

plot_data %>% 
  ggplot(aes(days_since_dst, weighted_mean_prop)) +
  geom_vline(xintercept = days(0), color = "grey60") +
  geom_point(aes(color = is_dark)) +
  geom_smooth(aes(group = is_dark), method = "lm", color = "grey40", se = FALSE) +
  scale_color_manual(values = c("#2b8cbe", "#feb24c")) +
  labs(
    title = "DST Discontinuity (20 minute period) over 9 cities",
    x = "Days since DST change",
    y = "Proportion minority drivers stopped - Mean over cities",
    color = NULL
  ) +
  facet_grid(season ~ .)
```

@@

discontinuity_states %>% 
  group_by(days_since_dst, season, is_dark, state) %>% 
  summarise(prop_minority = sum(is_minority_demographic) / n()) %>% 
  summarise(med_prop_minority = median(prop_minority)) %>% 
  ggplot(aes(days_since_dst, med_prop_minority)) +
  geom_vline(xintercept = days(0), color = "grey60") +
  geom_point(aes(color = is_dark)) +
  geom_smooth(aes(group = is_dark), se = FALSE, method = "lm", color = "grey20") +
  scale_color_manual(values = c("#2b8cbe", "#feb24c")) +
  facet_grid(. ~ season) +
  labs(
    title = "DST Discontinuity (20 minute period) over 10 states",
    x = "Days since DST change",
    y = "Proportion minority drivers stopped - Mean over states",
    color = NULL
  )
```

```{r}
city %>%

```

>>>>

```{r}
data <- read_rds(here::here("cache", "vod_dst_data.Rds"))
```

```{r}
min_minutes <-
  data %>% 
  group_by(geo_control) %>% 
  summarise(min_dusk = min(dusk_minute), max_sunset = max(sunset_minute))

data %>% 
  left_join(min_minutes, by = "geo_control") %>% 
  filter(minute >= min_dusk, minute <= max_sunset) %>% 
  count(geo_control, season = month(date) < 6) %>% 
  arrange(desc(n))
```

@@

```{r}
plot_prop_minority_by_time(
    filter(data, !year(date) %in% c(2007, 2008)),
    city_only = "Madison",
    title = "Madison, WI",
    min_clock_time = hms::hms(hours = 18, min = 30), 
    max_clock_time = hms::hms(hours = 19, min = 30),
    smooth_method = "lm"
  )
````

@@

```{r}
model.minute <- read_rds(here::here("cache", "model_minute_bin.rds"))
model.geo <- read_rds(here::here("cache", "model_geo_adj_minute_bin.rds"))
```

@@

by_city_dst %>%  
  filter(city != "El Paso") %>% 
  mutate(low = estimate - 1.96*std.error, high = estimate + 1.96*std.error) %>% 
  ggplot(aes(reorder(city, estimate), estimate)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_point() +
  geom_errorbar(aes(ymin = low, ymax = high)) +
  theme(
    axis.text.x = element_text(hjust = 1, angle = 30)
  ) +
  labs(
    x = NULL,
    y = "Coefficient on isDarkTRUE with 95% conf int",
    title = "Clock-time and Subgeography Adjusted DST Models by City"
  ) +
  facet_grid(season ~ .)
```

@@

plot_data <-
  full_data %>% 
  filter(
    days_since_dst < days(15),
    days_since_dst > days(-15),
    minute >= min_minute,
    minute <= min_minute + 20,
    city != "El Paso"
  ) %>% 
  mutate(
    is_dark = factor(is_dark, levels = c(TRUE, FALSE), labels = c("Dark", "Light"))
  ) %>% 
  group_by(days_since_dst, season, is_dark) %>% 
  summarise(prop_minority = sum(is_minority_demographic) / n(), n = n())

plot_data %>% 
  ggplot(aes(days_since_dst, prop_minority)) +
  geom_vline(xintercept = days(0), color = "grey60") +
  geom_point(aes(color = is_dark)) +
  geom_smooth(method = "lm", color = "grey40", se = FALSE) +
  scale_color_manual(values = c("#2b8cbe", "#feb24c")) +
  labs(
    title = "DST Discontinuity (20 minute period) over 9 cities",
    x = "Days since DST change",
    y = "Proportion minority drivers stopped - Mean over cities",
    color = NULL
  ) +
  facet_grid(season ~ .)
```

@@

discontinuity_states %>% 
  group_by(days_since_dst, season, is_dark) %>% 
  summarise(prop_minority = sum(is_minority_demographic) / n(), n = n()) %>% 
  ggplot(aes(days_since_dst, prop_minority)) +
  geom_vline(xintercept = days(0), color = "grey60") +
  geom_point(aes(color = is_dark)) +
  geom_smooth(se = FALSE, method = "lm", color = "grey20") +
  scale_color_manual(values = c("#2b8cbe", "#feb24c")) +
  facet_grid(. ~ season) +
  labs(
    title = "DST Discontinuity (20 minute period) over 10 states",
    x = "Days since DST change",
    y = "Proportion minority drivers stopped - Mean over states",
    color = NULL
  )
```

### 2018-12-13 lib/vod/vodReport.Rmd
cleaned up vod into lib/vod

---
title: "OPP Veil of Darkness Report"
author: "Samantha Robertson"
output: pdf_document
---

@@

# Abstract
In this report we implement the Veil of Darkness test (Grogger & Ridgeway, 2006) with a city-level dataset containing `r format(nrow(data.cities), big.mark = ",")` traffic stops across seven cities, and with a statewide dataset containing `r format(nrow(data.states), big.mark = ",")`  stops across ten states. The report is intended for readers who are relatively familiar with the test, so the explanation of its details is brief. Using logistic regression, we find statistically significant evidence of discrimination at the nationwide level in both datasets when controlling for the time and location of the stop. City-specific analysis identifies evidence of discrimination in Madison, WI; San Francisco, CA; Nashville, TN; and Philadelphia, PA. State-specific analysis __TODO__.

@@

[^2]: Data exploration has not been as rigorously performed on the statewide data. It is possible that some states or years should be removed due to data quality issues or very small percentages of minority drivers.

@@

In addition, we fit a second model that additionally controls for and geographic subdivision with an additive categorical variable. Since it is reasonable, if not likely, that discrimination varies greatly by city or state, we also fit this time and geography controlled model on each city and each state individually.

A concern raised by G & R is the role of seasonal effects. To control for seasonal effects we fit a third model on the full dataset with an additional additive control that treats the month of the year as a categorical variable. In addition, we fit the time and geography-controlled regression on subsets of the data that includes only the two weeks before and after the two annual Daylight Savings Time (DST) shifts.[^4]

[^4]: The model fit with a month control, and the DST models are only fit on the city-level data, due to time constraints and the large size of the aggregated statewide dataset, however the code is readily available to fit these models on the statewide data.

@@

G & R define the veil of darkness parameter $K_{vod} = \frac{P(B|S, d = 0)P(\bar{B} | S, d = 1)}{P(\bar{B}|S, d = 0)P(B | S, d = 1)}$, where $B$ is the event a driver is black, $S$ is the event a driver is stopped, and $d$ is darkness as we define it, where $d = 1$ indicates it is dark and $d = 0$ indicates it is light. Their analysis focuses on $log K$, which is negative in the presence of discrimination. As described in the paper, $-\beta_1$, where $\beta_1$ is the coefficient on the darkness variable in the logistic regression models, is an estimate for $logK$.  In this analysis we choose to consider the coefficient directly, as we find this a more intuitive and interpretable approach. A negative coefficient $\beta_1$ indicates that darkness is associated with a lower proportion of minorities stopped, and thus a negative coefficient is indicative of VOD discrimination.

@@

Figure 1 suggests that, between 6:30 and 7:30 PM, the proportion of minority drivers in the population of stopped drivers is lower after dark than before dark. The percentage of drivers stopped between 90 and 85 minutes before sunset is between 0.5 and 3 percentage points higher than the percentage of minority drivers amongst those stopped between 90 and 105 minutes after sunset.[^5]

[^5]: It should be noted that if we calculate the proportion of minorities by city (or geographic subdivision) and plot the mean of these proportions for each clock time and time relative to sunset, we produce a plot with very slightly positively sloped lines. This is likely due to the increased influence of smaller cities, which, as we see in Figure 2, tend to show less evidence of discrimination. We felt that Figure 1 is more representative of the effect present in the regressions included in Table 5, as it effectively weights the cities by their size, however, this phenomenon is worth further investigation.

@@

The lines in Figure 4 are sloping slightly downwards in the direction of the days where it is dark at the time of the stops. This suggests there may be VOD discrimination apparent in the data.[^6]

[^6]: See 5. The same argument applies in this case.

@@

The results of the logistic regression models controlling for clock time only and for clock time and geography are shown in Table 8. The coefficient on the darkness variable is negative in both models, and larger in magnitude in the model that controls additionally for geography. Both coefficient estimates are statistically significant at the 0.05 level, suggesting there is VOD discrimination present.

@@

#### TODO: State by state plot & maybe discontinuity

# Discussion
With the exception of the clock time and geography adjusted models fit on only the city data in the DST period, we see statistically significant evidence of VOD discrimination across the logistic regression models fit on both the city-level and the statewide data. When we consider discrimination at a city-level, we see statistically significant discrimination in Madison, San Francisco, Philadelphia and Nashville. At the state-level __TODO__.

The lack of statistical significance in the DST analysis when adjusting for geographic subdivision is likely indicative that the restriction of the data to only four weeks of the year significantly reduces the power of the test. The coefficient estimates in the models fit only on the DST shift data look qualitatively similar to those in the models fit on all of the data, and the model fit on all the data with a control for month showed no significant difference from those that do not control for month. This provides further reason to believe that the discrepancy in statistical significance is a symptom of the significant reduction in the size of the datasets. The quantity of data could be increased by including cities that do not have subdivision data available and running only the clock-time adjusted regression, or adjusting for geography using the city rather than the subdivision.[^7] It will also be interesting to see whether the DST analysis on the much larger statewide dataset has the same issues of power.

[^7]: See 1 for the list of these cities. See Appendix 1 for preliminary results of this analysis.

@@

My sincere thanks to Sharad for the opportunity to work on this project. It has been a steep but incredibly valuable learning curve that I am certain will serve me extremely well in the future. I have learned so much not only working with the largest dataset I've ever touched, but also being surrounded by such experienced and intelligent people who have been able to teach me how to ask the right questions and demand rigor. I would especially like to thank Amy and Dan for their generous support and patience throughout the quarter. Amy, in particular, has given so much of her time to answer my endless questions, help me through countless bugs and all without ever making me feel silly or like I'm wasting her time. She is an invaluable role model and I can't thank her enough.

@@

>>>>

---
title: "OPP Veil of Darkness Report"
author: "Samantha Robertson"
date: 12/12/2018
output: pdf_document
---

@@

# Abstract
In this report we implement the Veil of Darkness test (Grogger & Ridgeway, 2006) with a city-level dataset containing `r format(nrow(data.cities), big.mark = ",")` traffic stops across seven cities, and with a statewide dataset containing `r format(nrow(data.states), big.mark = ",")`  stops across ten states. The report is intended for readers who are relatively familiar with the test, so the explanation of its details is brief. Using logistic regression, we find statistically significant evidence of discrimination at the nationwide level in both datasets when controlling for the time and location of the stop. City-specific analysis identifies evidence of discrimination in Madison, WI; San Francisco, CA; Nashville, TN; and Philadelphia, PA.

@@

[^2]: Data exploration has not been as rigorously performed on the statewide data. It is possible that some states or years should be removed due to data quality issues or very small percentages of minority drivers. All statewide results should be considered preliminary.

@@

In addition, we fit a second model that additionally controls for and geographic subdivision with an additive categorical variable. Since it is reasonable, if not likely, that discrimination varies greatly by city or state, we also fit this time and geography controlled model on each city individually[^4].

[^4]: There were issues with the state-by-state analysis that only appeared at the very end of the quarter, so it is not included. However, this is certainly something valuable to look at. See Footnote 9 for more details.

A concern raised by G & R is the role of seasonal effects. To control for seasonal effects we fit a third model on the full dataset with an additional additive control that treats the month of the year as a categorical variable. In addition, we fit the time and geography-controlled regression on subsets of the data that includes only the two weeks before and after the two annual Daylight Savings Time (DST) shifts.[^5]

[^5]: The model fit with a month control, and the DST models are only fit on the city-level data, due to time constraints and the large size of the aggregated statewide dataset, however the code is available to fit these models on the statewide data.

@@

G & R define the veil of darkness parameter $K_{vod} = \frac{P(B|S, d = 0)P(\bar{B} | S, d = 1)}{P(\bar{B}|S, d = 0)P(B | S, d = 1)}$, where $B$ is the event a driver is black, $S$ is the event a driver is stopped, and $d$ is darkness as we define it, where $d = 1$ indicates it is dark and $d = 0$ indicates it is light. Their analysis focuses on $log K$, which is negative in the presence of discrimination. As described in the paper, $-\beta_1$, where $\beta_1$ is the coefficient on the darkness variable in the logistic regression models, is an estimate for $logK$.  In this analysis we choose to consider the coefficient directly, as we find this a more intuitive and interpretable approach. A negative coefficient estimate $\hat{\beta}_1$ indicates that darkness is associated with a lower proportion of minorities stopped, and thus a negative coefficient is indicative of VOD discrimination.

@@

Figure 1 suggests that, between 6:30 and 7:30 PM, the proportion of minority drivers in the population of stopped drivers is lower after dark than before dark. The percentage of drivers stopped between 90 and 85 minutes before sunset is between 0.5 and 3 percentage points higher than the percentage of minority drivers amongst those stopped between 90 and 105 minutes after sunset.[^6]

[^6]: It should be noted that if we calculate the proportion of minorities by city (or geographic subdivision) and plot the mean of these proportions for each clock time and time relative to sunset, we produce a plot with very slightly positively sloped lines. This is likely due to the increased influence of smaller cities, which, as we see in Figure 2, tend to show less evidence of discrimination. We felt that Figure 1 is more representative of the effect present in the regressions included in Table 5, as it effectively weights the cities by their size, however, this phenomenon is worth further investigation.

@@

The lines in Figure 4 are sloping slightly downwards in the direction of the days where it is dark at the time of the stops. This suggests there may be VOD discrimination apparent in the data.[^7]

[^7]: See Footnote 6. The same argument applies in this case.

@@

The results of the logistic regression models controlling for clock time only and for clock time and geography are shown in Table 8. The coefficient on the darkness variable is negative in both models, and larger in magnitude in the model that controls additionally for geography. Both coefficient estimates are statistically significant at the 0.05 level, suggesting there is VOD discrimination present.[^8]

[^8]: The state-by-state analysis, using the model with county controls, returned exactly the same coefficient and standard error on the darkness variable ($\hat{\beta}_1 = -0.0949$, $SE = 0.00406$) as the aggregated model. I can't find any obvious bugs with the code, which makes me concerned that there is an issue with the way the models are being fit. Since the dataset is so large, I didn't have time to investigate what's going on here before the end of the quarter. The code is in the `run_state_analysis` function in `run_vod.R` (see Appendix 2).

@@

# Discussion
With the exception of the clock time and geography adjusted models fit on only the city data in the DST period, we see statistically significant evidence of VOD discrimination across the logistic regression models fit on both the city-level and the statewide data. When we consider discrimination at a city-level, we see statistically significant discrimination in Madison, San Francisco, Philadelphia and Nashville.

The lack of statistical significance in the DST analysis when adjusting for geographic subdivision is likely indicative that the restriction of the data to only four weeks of the year significantly reduces the power of the test. The coefficient estimates in the models fit only on the DST shift data look qualitatively similar to those in the models fit on all of the data, and the model fit on all the data with a control for month showed no significant difference from those that do not control for month. This provides further reason to believe that the discrepancy in statistical significance is a symptom of the significant reduction in the size of the datasets. The quantity of data could be increased by including cities that do not have subdivision data available and running only the clock-time adjusted regression, or adjusting for geography using the city rather than the subdivision.[^9] It will also be interesting to see whether the DST analysis on the much larger statewide dataset has the same issues of power.

[^9]: See 1 for the list of these cities. See Appendix 1 for preliminary results of this analysis.

@@

My sincere thanks to Sharad for the opportunity to work on this project. It has been a steep but incredibly valuable learning curve. I am so grateful to have worked with such experienced and intelligent people who have been able to teach me how to ask the right questions and demand rigor. I would especially like to thank Amy and Dan for their generous support and patience throughout the quarter. Amy, in particular, has given so much of her time to answer my endless questions and help me through countless bugs, always with such kindness.

@@

# Works Cited

Grogger, Jeffrey and Greg Ridgeway, Testing for Racial Profiling in Traffic Stops From Behind a Veil of Darkness. American Statistical Association, 2006. https://www.rand.org/pubs/reprints/RP1253.html
